Basics of Hardware Cache Optimization

### 2017-06-22 tex/com.tex
Add some org-reveal slides; expand, revise, tweak
http://orgmode.org/
https://github.com/yjwen/org-reveal
https://github.com/hakimel/reveal.js/

% > In contrast to the external-memory model, algorithms in the cache-oblivious model
% > cannot explicitly manage the cache (issue block-read and block-write requests).  This
% > loss of freedom is necessary because the block and cache sizes are unknown.
%      -- http://erikdemaine.org/papers/BRICS2002/paper.pdf (erikcom), page 5
Cache misses take the place of \acrshortpl{io} \x{used in the \gls{emm}} because
cache-oblivious algorithms don't \alts{manage the cache, read or write cache lines}
explicitly.  This wouldn't be possible since the algorithms know neither the cache nor
the cache line size~\cite[5]{erikcom}.
% > The ideal cache uses the optimal off-line strategy of replacing the cache line whose
% > next access is furthest in the future
%      -- http://supertech.csail.mit.edu/papers/FrigoLePr99.pdf#page=1
Instead, the \gls{com} uses the optimal replacement strategy of evicting the cache line
that won't be accessed for the longest time in the future (Bélády's Algorithm).  This
\alts{
  is strangely \alts{at odds, out of touch, unrealistic} with real-world caches that don't
  know the future,
  {seems like out-of-touch, theoretical ivory-tower nonsense},
}.
% > The ideal-cache model makes the perhaps-questionable assumptions that there are only
% > two levels in the memory hierarchy, that memory is managed automatically by an optimal
% > cache-replacement strategy, and that the cache is fully associative.  We address these
% > assumptions in Section 6, showing that to a certain extent, these assumptions entail
% > no loss of generality.
%      -- http://supertech.csail.mit.edu/papers/FrigoLePr99.pdf#page=2
%
% > LRU and FIFO replacement do just as well as optimal replacement up to a constant
% > factor of memory transfers and up a constant factor wastage of the cache.
%      -- http://erikdemaine.org/papers/BRICS2002/paper.pdf#page=6 (erikcom)
%
% > The same argument extends to a variety of other replacement strategies.
%      -- http://supertech.csail.mit.edu/papers/Prokop99.pdf#page=46
%
% > We show that algorithms with regular complexity bounds (Equation (7.1)) (including all
% > algorithms heretofore presented) can be ported to less-ideal caches incorporating
% > least-recently-used (LRU) or first-in, first-out (FIFO) replacement policies [24, p.
% > 378].
%      -- http://supertech.csail.mit.edu/papers/Prokop99.pdf#page=51
%
% > [I]n many cases [the COM] is provably within a constant factor of a more realistic
% > cache's performance.
%      -- https://en.wikipedia.org/wiki/Cache-oblivious_algorithm
%
% > [A]s long as the number of memory transfers depends polynomially on the cache size M,
% > then halving M will only affect the running time by a constant factor.
%      -- http://erikdemaine.org/papers/BRICS2002/paper.pdf#page=6 (erikcom)
%
% > Intuitively, algorithms that slow down by a constant factor when memory (M) is reduced
% > to half, are called regular.
%      -- Algorithms for Memory Hierarchies, page 196
However, \citeauthor{coa-thesis} proves that for many algorithms\footnote{%
   Those satisfying equation (7.1) in~\cite[46]{coa-thesis}.  If the number of cache
   misses incurred by the algorithm only \say{depends polynomially on the cache size
   \(M\)}, the \alts{equation, condition} is satisfied~\cite[6]{erikcom}.
   % Halving the cache size only increases the number of cache misses by a constant
   % factor?
}
it only increases cache misses by a constant factor compared to various feasible
replacement strategies%
% ~\cite[lemma 12, \pno~10]{coa-paper}.%
% ~\cite[corollary 13, \pno~10]{coa-paper}.%
~\cite[corollary 19, \pno~46]{coa-thesis}.%
% See <http://supertech.csail.mit.edu/papers/Prokop99.pdf#page=46>.
\footnote{%
   E.g. LRU, FIFO, and random replacement
}

@@

>>>>

% > In contrast to the external-memory model, algorithms in the cache-oblivious model
% > cannot explicitly manage the cache (issue block-read and block-write requests).  This
% > loss of freedom is necessary because the block and cache sizes are unknown.
%      -- http://erikdemaine.org/papers/BRICS2002/paper.pdf (erikcom), page 5
Cache misses take the place of \acrshortpl{io} \x{used in the \gls{emm}} because
cache-oblivious algorithms don't \alts{manage the cache, read or write cache lines}
explicitly.  This wouldn't be possible since the algorithms know neither the cache nor
the cache line size~\cite[5]{erikcom}.
% > The ideal cache uses the optimal off-line strategy of replacing the cache line whose
% > next access is furthest in the future
%      -- http://supertech.csail.mit.edu/papers/FrigoLePr99.pdf#page=1
Instead, the \gls{com} uses the optimal replacement strategy of evicting the cache line
that won't be accessed for the longest time in the future (Bélády's Algorithm).  This
\alts{
  is strangely \alts{at odds, out of touch, unrealistic} with real-world caches that don't
  know the future,
  {seems like out-of-touch, theoretical ivory-tower nonsense},
}.
% > The ideal-cache model makes the perhaps-questionable assumptions that there are only
% > two levels in the memory hierarchy, that memory is managed automatically by an optimal
% > cache-replacement strategy, and that the cache is fully associative.  We address these
% > assumptions in Section 6, showing that to a certain extent, these assumptions entail
% > no loss of generality.
%      -- http://supertech.csail.mit.edu/papers/FrigoLePr99.pdf#page=2
%
% > LRU and FIFO replacement do just as well as optimal replacement up to a constant
% > factor of memory transfers and up a constant factor wastage of the cache.
%      -- http://erikdemaine.org/papers/BRICS2002/paper.pdf#page=6 (erikcom)
%
% > The same argument extends to a variety of other replacement strategies.
%      -- http://supertech.csail.mit.edu/papers/Prokop99.pdf#page=46
%
% > We show that algorithms with regular complexity bounds (Equation (7.1)) (including all
% > algorithms heretofore presented) can be ported to less-ideal caches incorporating
% > least-recently-used (LRU) or first-in, first-out (FIFO) replacement policies [24, p.
% > 378].
%      -- http://supertech.csail.mit.edu/papers/Prokop99.pdf#page=51
%
% > [I]n many cases [the COM] is provably within a constant factor of a more realistic
% > cache's performance.
%      -- https://en.wikipedia.org/wiki/Cache-oblivious_algorithm
%
% > [A]s long as the number of memory transfers depends polynomially on the cache size M,
% > then halving M will only affect the running time by a constant factor.
%      -- http://erikdemaine.org/papers/BRICS2002/paper.pdf#page=6 (erikcom)
%
% > Intuitively, algorithms that slow down by a constant factor when memory (M) is reduced
% > to half, are called regular.
%      -- Algorithms for Memory Hierarchies, page 196
However, \citeauthor{coa-thesis} proves that for many algorithms\footnote{%
   Those satisfying equation (7.1) in~\cite[46]{coa-thesis}.  If the number of cache
   misses incurred by the algorithm only \say{depends polynomially on the cache size
   \(M\)}, the \alts{equation, condition} is satisfied~\cite[6]{erikcom}.
   % Halving the cache size only increases the number of cache misses by a constant
   % factor?
}
it only increases cache misses by a constant factor compared to various feasible
replacement strategies%
% ~\cite[lemma 12, \pno~10]{coa-paper}.%
% ~\cite[corollary 13, \pno~10]{coa-paper}.%
~\cite[corollary 19, \pno~46]{coa-thesis}.%
% See <http://supertech.csail.mit.edu/papers/Prokop99.pdf#page=46>.
\footnote{%
   e.g. LRU, FIFO, and random replacement
}

@@

The straightforward way to transpose \alts{
   an \(m\times n\) matrix \(D\),
   a matrix \(D\in\mathbb{R}^{m\times{}n}\),
}
out-of-place is to use two loops like so:
\begin{minted}[autogobble]{c}
   for (int i = 0; i < m; ++i)
      for (int j = 0; j < n; ++j)
         E[j][i] = D[i][j];
\end{minted}
Assuming \(D\) and \(E\) are stored in row-major layout (as they would be in the C and
\cpp{} languages), the reads from \(D\) are sequential memory accesses but the writes to
\(E\) are not.

\alts{When \(D^\mathsf{T}\) has, For} sufficiently long rows (\(m > B\))\footnote{%
   \(D^\mathsf{T}\) will be an \(n\times m\) matrix, so \(m\) is its number of columns.
},
every consecutive access will be to a different cache line.  If it has sufficiently many
rows, no access will be to a cache line that is still loaded; once the inner loop
completes, the cache line holding \alts[2]{\mintinline{c}{E[0][i]}, \(B_{0,i}\)} will have
been evicted.%
\footnote{%
   assuming LRU or FIFO replacement % FIXME: but we use optimal replacement!
}
Therefore, this algorithm incurs \(\Theta(mn)\) cache misses.

The algorithm is \alts{by definition, technically} already cache-oblivious since it
doesn't use \(M\) or \(B\), but it is not optimal\x{ in the \gls{com}}.  We can do better
with a divide-and-conquer approach.
% The code may only be 72 characters wide (75 when counting the indentation that will be
% stripped).  Note that the array arguments will decay to pointers.
\begin{minted}[autogobble, mathescape]{c}
   // Transpose the submatrix $(d_{ij})_{i\in I,\:j\in J}$.
   void transpose(int I[2], int J[2], int D[m][n], int E[n][m]) {
      int num_rows = 1 + I[1] - I[0];
      int num_cols = 1 + J[1] - J[0];
      if (num_cols == 1 && num_rows == 1) {
         E[J[0]][I[0]] = D[I[0]][J[0]];
      } else if (num_cols <= num_rows) {
         // Horizontally slice D into two submatrices and recurse.
         transpose((int[2]){I[0], I[0] + num_rows / 2 - 1}, J, D, E);
         transpose((int[2]){I[0] + num_rows / 2, I[1]}, J, D, E);
      } else { /* Vertically slice D analogously... */ }
   }
\end{minted}
% // Transpose the submatrix $(d_{ij}),\: i\in I, j\in J$.
% // Transpose the submatrix $(d_{ij})_{I_0<i<I_1, J_0<j<J_1}$
% // Transpose the submatrix $D\left[...\right]$

\begin{figure}
   \centering
   \begin{tikzpicture}
      \datavisualization
      [scientific axes=clean,
       x axis={logarithmic,
               label={Matrix Size (\si{\mebi\byte})}, length=0.8\textwidth, ticks={
                  major at={1, 2, 4, 8, 16, 32, 64, 128, 256},
               },
              },
       y axis={include value=0, grid=at ticks, label={Speedup}, length=6cm,
         % ticks={major also at={1}}
       },
       visualize as scatter,
       scatter={style={mark=*, mark options={scale=.65}}}]
         data [read from file=xpose/speedup.csv, separator=\space];
   \end{tikzpicture}
   \caption{TODO}
   \label{fig:TODO}
\end{figure}

% > A cache-oblivious algorithm is simply a RAM algorithm; it is only the analysis that
% > differs.
%      -- https://pdfs.semanticscholar.org/2f5a/76ccdc71971b746fbe7ff54db86c65a73e91.pdf
%
% Algorithms are cache-oblivious by default.

### 2017-06-22 tex/list-vs-vector.tex
Add some org-reveal slides; expand, revise, tweak
http://orgmode.org/
https://github.com/yjwen/org-reveal
https://github.com/hakimel/reveal.js/

\alts{%
   The C++ program shown in \cref{lst:ithare},
   The following C++ program,
}%
, adapted from \textcite{bigos}, initializes a number of \acrshort{stl} containers
with random numbers and measures the processor time needed to \alts{sum all of them,
compute the sum}.
%
I first ran it with \mintinline{cpp}{Container} being a type alias for
\mintinline{cpp}{std::list}, then for \mintinline{cpp}{std::vector}.%
\footnote{%
   Each compiled with \x{the \gls{gcc}} and \mintinline{text}{-O3} and
   \mintinline{text}{-march=native}
}
Either way, the asymptotic complexity is \(\Theta(N)\).

@@

\subsection{``True'' OO Style}

@@

In object-oriented systems, objects are typically referred to by pointers to a common base
class.  A polymorphic container of such pointers allows for dynamic dispatch of virtual
functions.  However, this carries the risk of degrading the \x{superior} performance of a
sequential data structure to that of a list%
% ~\cite[\href{https://youtu.be/0iWb_qi2-uI?t=51m22s}{51:22}]{stroustrup-talk}.
~\cite[51:22]{stroustrup-talk}.
See \cref{fig:true-oo}.

\begin{figure}
   \centering
   \input{tex/oo-picture}
   \captionsetup{width=.9\linewidth} % FIXME: hacks!
   \caption[Array of Pointers]{Array of Pointers; while the array is compact, the actual
   objects may not be laid out sequentially in memory.}
   \label{fig:true-oo}
\end{figure}

\subsection{Unrolled List}

>>>>

\alts{%
   The \cpp{} program shown in \cref{lst:ithare},
   The following \cpp{} program,
}%
, adapted from \textcite{bigos}, initializes a number of \acrshort{stl} containers
with random numbers and measures the processor time needed to \alts{sum all of them,
compute the sum}.
%
I first ran it with \mintinline{cpp}{Container} being a type alias for
\mintinline{cpp}{std::list}, then for \mintinline{cpp}{std::vector}.%
\footnote{%
   Each compiled with \x{the \gls{gcc}} and \mintinline{text}{-O3} and
   \mintinline{text}{-march=native}
}
Either way, the asymptotic complexity is \(\Theta(N)\).

@@

\subsection{Notes}

\subsubsection{``True'' OO Style}

@@

In \gls{oo} systems, variables are typically referred to by pointers to a common base
class.  A polymorphic container of such pointers allows for dynamic dispatch of virtual
functions.  However, this carries the risk of degrading the \x{superior} performance of a
sequential data structure to that of a list%
% ~\cite[\href{https://youtu.be/0iWb_qi2-uI?t=51m22s}{51:22}]{stroustrup-talk}.
~\cite[51:22]{stroustrup-talk}.
See \cref{fig:true-oo}.

\begin{figure}
   \centering
   \input{tex/oo-picture}
   \captionsetup{width=.9\linewidth} % FIXME: hacks!
   \caption[Array of Pointers]{Array of Pointers; the array is compact but the actual
   objects may \alts{be scattered across memory pretty randomly, not be laid out
   sequentially in memory}.}
   \label{fig:true-oo}
\end{figure}

\subsubsection{Unrolled List}
